\name{}: machine learning over interaction networks and ontology-based background knowledge predicts drug--target interactions

### 2021-04-21 paper/main.tex
Update download scripts

\abstract{\textbf{Motivation:} \textit{In silico} drug--target
  interaction (DTI) prediction is important for drug discovery and
  drug repurposing.  Approaches to predict DTIs can proceed
  indirectly, top-down, using phenotypic effects of drugs to identify
  potential drug targets, or they can be direct, bottom-up and use
  molecular information to directly predict binding potentials.  Both
  approaches can be combined with information about interaction
  networks. \\
  \textbf{Results:} We developed \name{} as a computational method
  that combines molecular features and ontology-encoded phenotypic
  effects of drugs with protein--protein interaction networks, and
  uses a graph neural networks to predict DTIs.  We demonstrate that
  drug effect features can exploit information in the interaction
  network whereas molecular features do not.  \name{} is designed to
  predict candidate drugs for a given protein; we use this formulation
  to show that common DTI datasets contain intrinsic biases with major
  affects on performance evaluation and comparison of DTI prediction
  methods. Using a modified evaluation scheme, we demonstrate that
  \name{} improves substantially over state of the art DTI prediction
  methods. \\
  \textbf{Availability:} \name{} source code and data necessary to
  reproduce results are freely available at
  \url{https://github.com/THinnerichs/DTI-VOODOO}. \\
  \textbf{Contact:}
  \href{tilman.hinnerichs@kaust.edu.sa}{tilman.hinnerichs@kaust.edu.sa}
  \\
  \textbf{Supplementary information:}Supplementary data are available
  at \textit{Bioinformatics} online.}

@@

As we want to learn from the similarity of drug side effects and
protein phenotypes, we use a deep half-twin neural network with a
contrastive loss using cosine similarity.  A half-twin neural network
aims to learn a similarity between two embeddings of variable but same
dimension. As the original feature space may have varying dimensionality, we first process them using a fully connected
neural network layer which takes as input an embedding and outputs a
representation of a particular size, i.e., we use this layer as a
trainable feature transformation and apply it to reduce the
representation size of the embeddings for drugs and proteins
separately.  An example structure for both types of features can be
found in Figure~\ref{fig:HalfTwinNetwork}.
% While a regular deep neural network, denoted by LFT, for feature space reduction is not particularly novel, we
The use of this trainable feature transformation layer enables
flexible experimentation as both ontology and molecular feature for
both drugs and proteins are reduced to the same dimensionality for
varying sizes of inputs; this allows for a high amount of modularity
across different experimental setups by adding different kinds of
features into the model. Additionally, the generated features may be
used for other tasks. % ; for
% example, the ontology LFT can be reused for a variety of DL2vec-based
% features with respect to other ontologies and hypotheses
We follow the results of \textit{DL2vec} \cite{DL2vec2020} and use as
activation function $\sigma := \mathrm{LeakyReLU}$ which leads to
improved performance compared to other activation functions.

>>>>

\abstract{\textbf{Motivation:} \textit{In silico} drug--target
  interaction (DTI) prediction is important for drug discovery and
  drug repurposing.  Approaches to predict DTIs can proceed
  indirectly, top-down, using phenotypic effects of drugs to identify
  potential drug targets, or they can be direct, bottom-up and use
  molecular information to directly predict binding potentials.  Both
  approaches can be combined with information about interaction
  networks. \\
  \textbf{Results:} We developed \name{} as a computational method
  that combines molecular features and ontology-encoded phenotypic
  effects of drugs with protein--protein interaction networks, and
  uses a graph convolutional neural network to predict DTIs.  We demonstrate that
  drug effect features can exploit information in the interaction
  network whereas molecular features do not.  \name{} is designed to
  predict candidate drugs for a given protein; we use this formulation
  to show that common DTI datasets contain intrinsic biases with major
  affects on performance evaluation and comparison of DTI prediction
  methods. Using a modified evaluation scheme, we demonstrate that
  \name{} improves substantially over state of the art DTI prediction
  methods. \\
  \textbf{Availability:} \name{} source code and data necessary to
  reproduce results are freely available at
  \url{https://github.com/THinnerichs/DTI-VOODOO}. \\
  \textbf{Contact:}
  \href{tilman.hinnerichs@kaust.edu.sa}{tilman.hinnerichs@kaust.edu.sa}
  \\
  \textbf{Supplementary information:}Supplementary data are available
  at \textit{Bioinformatics} online.}

@@

As we want to learn from the similarity of drug side effects and
protein phenotypes, we use a deep half-twin neural network with a
contrastive loss using cosine similarity.  A half-twin neural network
aims to learn a similarity between two embeddings of variable but same
dimension. As the original feature space may have varying dimensionality, we first process them using a fully connected
neural network layer which takes as input an embedding and outputs a
representation of a particular size, i.e., we use this layer as a
trainable feature transformation and apply it to reduce the
representation size of the embeddings for drugs and proteins
separately.  An example structure for both types of features can be
found in Figure~\ref{fig:HalfTwinNetwork}.
% While a regular deep neural network, denoted by LFT, for feature space reduction is not particularly novel, we
The use of this trainable feature transformation layer enables
flexible experimentation as both ontology and molecular feature for
both drugs and proteins are reduced to the same dimensionality for
varying sizes of inputs; this allows for a high amount of modularity
across different experimental setups by adding different kinds of
features into the model. Additionally, the generated features may be
used for other tasks. % ; for
% example, the ontology LFT can be reused for a variety of DL2vec-based
% features with respect to other ontologies and hypotheses
We follow the results of \textit{DL2vec} \cite{DL2vec2020} and use $\sigma := \mathrm{LeakyReLU}$ as
activation function  which leads to
improved performance compared to other activation functions.
